---
output:
  pdf_document:
    highlight: default
urlcolor: blue
geometry:
- margin=1in
- paperheight=6.5in
- paperwidth=8.5in
header-includes:
- \usepackage{fancyhdr}
- \fancyhead{}
- \fancyfoot{}
- \chead{Visualization}
- \lhead{Lecture Notes for 46-921}
- \rhead{\footnotesize Slide \thepage}
- \pagestyle{fancy}
- \parindent=0in
- \renewcommand{\baselinestretch}{1.3}
- \usepackage{palatino}
- \usepackage{color}
- \usepackage [pagewise, mathlines, displaymath]{lineno}
- \input defs.tex
- \parskip 0.15in
fontsize: 11pt
#fontfamily: mathpazo
---

```{r setup, include=FALSE}
library(knitr)
knitr::opts_chunk$set(echo = TRUE, comment="", prompt=TRUE)
hook1 <- function(x){ gsub("```\n*```r*\n*", "", x) }
hook2 <- function(x){ gsub("```\n+```\n", "", x) }
knit_hooks$set(document = hook2)
setwd("~/Teaching/MSCF/DScourses/LectureNotes/Part5_Visualization")
```


\thispagestyle{fancy}
\linenumbers
\Large

\Huge
{\bf Part 5: Visualization}
\Large

#Summary

\vspace{-.2in}
This Part covers tools for visualizing data in R.

Powerful visualization techniques are critical to discovering and
understanding the relationships between variables.


\newpage

#Example Data Set

\vspace{-.2in}
First, we will revisit the data set we considered in the previous
Part, and the steps needed to properly format the data:

\small
```{r}
fulldata = read.table("q1_2017_all.csv", sep=",", header=T, quote="")
fulldata$Date = as.Date(as.character(fulldata$Date), format="%Y%m%d")
fulldata = fulldata[!duplicated(fulldata[,c(1,3)], fromLast=TRUE),]
fulldata$McapRank = factor(fulldata$McapRank, ordered=TRUE)
fulldata$TurnRank = factor(fulldata$TurnRank, ordered=TRUE)
fulldata$VolatilityRank = factor(fulldata$VolatilityRank, ordered=TRUE)
fulldata$PriceRank = factor(fulldata$PriceRank, ordered=TRUE)
```
\Large

\vspace{-.1in}
The entire data set is a bit large to work with conveniently, so
we are going to restrict consideration to 1000 equities:

\vspace{-.1in}
```{r}
tickkeep = unique(fulldata$Ticker)[seq(1,
      length(unique(fulldata$Ticker)),length=100)]
fulldata = fulldata[fulldata$Ticker %in% tickkeep,]
```


\newpage
# Classic Plotting Functions in R

\vspace{-.2in}
Prior to moving on the more advanced visualization functions,
it is important to understand the basic plotting capabilities of R.

We saw the use of `boxplot()` in a previous Part. A similar view of
the shape of a distribution can be obtained using a \textcolor{red}{histogram},
formed using the function `hist()`:

```{r,eval=FALSE}
hist(log10(fulldata$Hidden), 
     xlab="Log Number of Hidden Trades")
```
The title on the plot can be changed using the `main` argument.

The number of bins in the histogram can be altered using the `breaks` argument.

```{r,echo=FALSE,fig.align="center"}
hist(log10(fulldata$Hidden),  xlab="Log Number of Hidden Trades")
```

\newpage
\textcolor{red}{Scatter plots} are ubiquitous, as well. They are created simply
in R using the `plot()` function:

```{r, fig.align="center", fig.height=5, eval=FALSE}
plot(log10(fulldata$Hidden),
     log10(fulldata$TradesForHidden),
     xlab="Log Number of Hidden Trades", 
     ylab="Log Total Number of Trades", pch=".")
```
The argument `pch` changes the plotting character. In a case such as this,
with so many dots on the plot, the default plotting character is too large.

Many other properties of the scatter plot can be adjusted using the `par()`
function. See `help(par)` for more information.

```{r, fig.align="center", fig.height=5, echo=FALSE}
plot(log10(fulldata$Hidden),
     log10(fulldata$TradesForHidden),
     xlab="Log Number of Hidden Trades", 
     ylab="Log Total Number of Trades", pch=".")
```

\newpage
#Introduction to ggplot2

\vspace{-.2in}
The package `ggplot2` contains a great deal of valuable 
data visualization tools. This package can be installed,
along with other useful packages for working with large and
complex data sets, via the `tidyverse`
meta-package:

```{r,eval=FALSE}
install.packages("tidyverse")
```

The collection of packages in tidyverse comprise ``a collection of
R packages that share common philosophies and are designed to
work together." -- tidyverse.org


\newpage
The `ggplot2` package implements a \textcolor{red}{grammar of graphics},
where plots are created by combining \textcolor{red}{geoms}.

The syntax is a little awkward at first, but this approach yields a
great deal of flexibility, and a great deal of power in exploring
features in a data set.

A plot is initialized by using the `ggplot()` function:

```{r,include=FALSE}
library(ggplot2)
```

```{r,eval=FALSE}
ggplot(data = fulldata)
```

\newpage

To the basic plot, one adds the "aesthetic" via the
`mapping` option. An aesthetic a generic
concept for the visual features of a plot. In this case,
we are specifying that the variable on the horizontal
axis is `Hidden`, and the variable on the vertical
axis is `Cancels`:

```{r, eval=FALSE}
ggplot(data = fulldata, mapping = 
               aes(x=Hidden, y=Cancels))
```

We can now add onto this base to create plots with
complex features. You can, for instance, define
```{r, eval=FALSE}
baseplot = ggplot(data = fulldata, mapping = 
               aes(x=Hidden, y=Cancels))
```
and add onto `baseplot` in what follows.

\newpage
#A Simple Scatter Plot

\vspace{-.2in}
Consider the following syntax:

```{r, eval=FALSE}
baseplot + geom_point()
```

Note how the `geom_point()` function is used to add
a scatter plot onto the current figure, as stored
in `baseplot`.

\newpage
```{r, fig.height=4,fig.width=6,fig.align='center',echo=FALSE}
baseplot = ggplot(data = fulldata, mapping = 
               aes(x=Hidden, y=Cancels))
baseplot + geom_point()
```

\newpage
Now we will work to improve the appearance of this plot.

```{r, fig.height=4,fig.width=6,fig.align='center'}
baseplot + 
  geom_point(size=0.3, color="blue") + 
  labs(x="Trades on Hidden Orders", 
      y="Cancelled Orders",
      title="Cancelled Orders versus Hidden Orders",
      subtitle="Daily, First Quarter of 2017")
```

\newpage
__Exercise:__ Describe the changes that were made to improve the plot.

\answerlines{8}

\newpage
Next, we will convert the axes to the log scale:

```{r, fig.height=4,fig.width=6,fig.align='center',warning=FALSE}
baseplot + 
  geom_point(size=0.3, color="blue") + 
  labs(x="Trades on Hidden Orders", 
     y="Cancelled Orders",
     title="Cancelled Orders versus Hidden Orders",
     subtitle="Daily, First Quarter of 2017") + 
  scale_x_log10() + scale_y_log10()
```

\newpage
The `color` argument can be switched to vary with the levels of one
of the factors.

```{r, fig.height=4,fig.width=7,fig.align='center',warning=FALSE}
baseplot + 
 geom_point(size=0.3, mapping=aes(color=Security)) + 
 labs(x="Trades on Hidden Orders", 
     y="Cancelled Orders",
     title="Cancelled Orders versus Hidden Orders",
     subtitle="Daily, First Quarter of 2017") + 
 scale_x_log10() + scale_y_log10()
```

\newpage

#Bar Plots

\vspace{-.2in}

Basic bar charts can be created to inspect the distribution
of a factor, using the `geom_bar()` function:

```{r, fig.align="center", fig.height=4}
ggplot(data=fulldata, 
    mapping=aes(x=Security, fill=Security)) +
  geom_bar() +
  labs(x="Security Type", 
       y="Total Daily Observations")
```

\newpage
__Exercise:__ Now that we have seen a few examples, try to explain
the notion of the "aesthetic."

\answerlines{8}

\newpage

For the next plot, we will restrict to data that do not have missing
values on `McapRank`, and are of type "Stock":

```{r,include=FALSE}
library(dplyr)
```

```{r}
fulldataStock = filter(fulldata, 
      Security == "Stock" & !is.na(McapRank))
```

We will adjust the aesthetic so that the "fill" color
of the bars corresponds to the level of `VolatilityRank`:

```{r, fig.align="center", fig.height=4}
ggplot(data=fulldataStock, 
    mapping=aes(x=McapRank, fill=VolatilityRank)) +
  geom_bar() +
  labs(x="Market Capitalization Rank", y="Count", 
    fill="Volatility Rank")
```

\newpage
The bars in the previous plot are not of equal height because of
the subsampling we did to reduce to 1000 equities.

We can change vertical axis to proportion by setting
`position="fill"`:
```{r, fig.align="center", fig.height=4}
ggplot(data=fulldataStock, 
    mapping=aes(x=McapRank, fill=VolatilityRank)) +
  geom_bar(position="fill") +
  labs(x="Market Capitalization Rank", 
    y="Proportion", fill="Volatility Rank")
```

\newpage
By setting
`position="dodge"`, the bars are arranged more like a histogram:
  
```{r, fig.align="center", fig.height=4}
ggplot(data=fulldataStock, 
    mapping=aes(x=McapRank, fill=VolatilityRank)) +
  geom_bar(position="dodge") +
  labs(x="Market Capitalization Rank", 
    y="Count", fill="Volatility Rank")
```

\newpage
__Exercise:__ Why would you not want to create a scatter plot to explore
the relationship between Market Capitalization Rank and Volatility Rank?
Consider how the function `geom_jitter()` can help in this regard.

\answerlines{8}

```{r, eval=FALSE, echo=FALSE}
ggplot(data=fulldata[complete.cases(fulldata) & fulldata$Security=="Stock",]) +
  geom_jitter(mapping=aes(x=McapRank,y=VolatilityRank),width=0.15, height=0.15,size=0.01) +
  labs(x="Market Capitalization Rank", y="Volatility Rank")
```

\newpage

#Histograms

\vspace{-.2in}
The function `geom_histogram()` will create a histogram:

```{r, fig.align="center",fig.height=2.25,fig.width=4, warning=FALSE}
ggplot(data=fulldata, mapping=aes(x=Hidden)) + 
  geom_histogram(binwidth=0.5, fill="blue") + 
  scale_x_log10() +
  labs(x="Trades on Hidden Orders", y="Count")
```

\newpage
Overlapping histograms can be difficult to interpret (try it!).
An alternative is to create \textcolor{red}{frequency polygons}
using `geom_freqpoly()`:

```{r, fig.align="center",fig.height=2.0,fig.width=4, warning=FALSE}
ggplot(data=fulldata,
       mapping=aes(x=Hidden, color=Security)) + 
  geom_freqpoly(binwidth=0.5, fill="blue") + 
  scale_x_log10() +
  labs(x="Trades on Hidden Orders", y="Count")
```

\newpage
__Exercise:__ Both `geom_histogram()` and `geom_freqpoly()` have an argument
`binwidth`. Discuss the technical and practical effects of varying `binwidth`.

\answerlines{8}

\newpage
#Boxplots

\vspace{-.2in}
\textcolor{red}{Side-by-side boxplots} are also useful for comparing the distributions
of variables over different values of a factor. For example,

```{r, fig.align="center",fig.height=4.0, warning=FALSE}
ggplot(data=fulldataStock, 
       mapping=aes(x=VolatilityRank, y=Hidden)) +
  geom_boxplot() + scale_y_log10() +
  labs(y="Trades on Hidden Orders", 
       x="Volatility Rank")
```

\newpage
A \textcolor{red}{violin plot} shows the same type of information, with
a little more detail. The width of a "violins" is greater in areas where
a greater proportion of the observations lie.

```{r, fig.align="center",fig.height=4.0, warning=FALSE}
ggplot(data=fulldataStock, 
        mapping=aes(x=VolatilityRank, y=Hidden)) +
  geom_violin() + scale_y_log10() +
  labs(y="Trades on Hidden Orders", 
       x="Volatility Rank")
```


\newpage
#Facets

\vspace{-.2in}
\textcolor{red}{Facets} refer to an arrangment of plots on which
one or more factors vary. For example, return to our scatter
plot example above, and use the `facet_grid()` function to break
up the plots by month:

```{r, fig.height=4,fig.width=7,fig.align='center',warning=FALSE}
baseplot + 
 geom_point(size=0.3, mapping=aes(color=Security)) + 
 labs(x="Trades on Hidden Orders", 
     y="Cancelled Orders",
     title="Cancelled Orders versus Hidden Orders",
     subtitle="Daily, First Quarter of 2017") + 
 scale_x_log10() + scale_y_log10() +
 facet_grid(.~factor(months(Date),
      levels=c("January","February","March")))
```

\newpage
__Exercise:__ What happens in the previous example if the command
```{r,eval=FALSE}
facet_grid(.~months(Date))
```
is used instead? What about if use the following?
```{r,eval=FALSE}
facet_grid(months(Date)~.)
```

\answerlines{6}


\newpage
Two factors can vary:
```{r, fig.height=5,fig.width=7,fig.align='center',warning=FALSE}
baseplot + 
 geom_point(size=0.3, color="red") + 
 labs(x="Trades on Hidden Orders", 
     y="Cancelled Orders",
     title="Cancelled Orders versus Hidden Orders",
     subtitle="Daily, First Quarter of 2017") + 
 scale_x_log10() + scale_y_log10() +
 facet_grid(Security~factor(months(Date),
      levels=c("January","February","March")))
```
\newpage
#Time Series Plots

\vspace{-.2in}
We will, of course, be working with a great deal of \textcolor{red}{time
series data}. The `geom_line()` function is useful:

```{r, fig.height=5,fig.width=7,fig.align='center',warning=FALSE}
ggplot(data=filter(fulldata,Ticker=="AMD"), 
       mapping=aes(x=Date,y=Hidden)) +
  geom_line(color="blue") +
  labs(x="Date", y="Trades on Hidden Orders", 
     title="Change in Hidden Orders over Time",
     subtitle="AMD, daily, first quarter of 2017")
```

